# Posterior Predictive Meta-Analytic Network

Ecological information is accumulated across numerous independent studies that describe interactions among biotic and abiotic components of ecosystems. Although this literature contains a wealth of ecological information, it is dispersed across publications, expressed using diverse statistical measures, and often communicated qualitatively. Consequently, ecological understanding remains fragmented rather than synthesized into coherent quantitative networks.

This fragmentation is unavoidable. Ecological systems are high-dimensional, and no individual study can simultaneously measure all relevant abiotic and biotic variables. Furthermore, ecological understanding develops as new mechanisms and interactions are better understood. As a result, ecological networks are necessarily incomplete and continuously updated from scattered information.

Existing syntheses frequently describe interactions only as positive or negative relationships. While such qualitative descriptions are useful for conceptual understanding, they provide little information about the magnitude of ecological responses. Conversely, conventional meta-analyses summarize information using standardized effect sizes, such as Cohen's d,

$$
d=\frac{\overline{x}_{treatment}-\overline{x}_{control}}{SD_{pooled}}
$$

while the enable comparison among study, they remove the original measurement units. Consequently, these effect sizes are difficult to interpret directly in ecological prediction because identical standardized effects can correspond to very different absolute ecological changes (Baguley, 2009; Tukey, 1969).

For predictive ecological networks, quantities expressed in natural units are often more informative. For example, estimating the change in EPT abundance associated with a 1 $mg\cdot L^{-1}$  decline in dissolved oxygen yields a relationship that can be directly applied to field observations, propagated through ecological networks, and evaluated against future measurements. Such quantities possess immediate operational meaning for prediction and management.

However, no general framework currently exists for systematically combining these heterogeneous quantitative relationships into an evolving ecological network. We therefore introduce the Posterior Predictive Meta-analytic Network (PPMN), a modular framework that synthesizes dispersed ecological information into a coherent predictive network. The PPMN combines Bayesian meta-analysis, probabilistic simulation, network modelling, and sequential updating to estimate and propagate ecological relationships without requiring a complete multivariate dataset. The framework supports prediction, structural evaluation, scenario exploration, and continual refinement as new information becomes available. In this way, qualitative ecological understanding is progressively formalized into an updateable quantitative network.

## Building blocks

A PPMN describes an ecological system as a generalized directed Bayesian network. *Vertices* (nodes), represent *ecological variables*, whereas directed *edges* (links), are represented by parameterized *edge-functions*. Each edge-function contains one or more *model parameters*. Information about these parameters is represented by probability distributions rather than by single fixed values. Together, the network structure, edge-functions, and parameter distributions determine the predictive behavior of the PPMN.

The figure below illustrates a simple PPMN containing four variables and three parameterized edge-functions. For each edge-function, information from prior distributions and study-derived estimates is synthesized into a posterior parameter distribution.

```{r, fig1_ppmn,  fig.width=8, fig.height=4, echo=F, warning=FALSE}

library(ggplot2)
library(grid)
library(gridExtra)
library(bezier)
library(EcoPostView)

data <- data.frame(
  x = c(2, 2, .5, 3.5),
  y = c(4, 2, 0.25, 0.25),
  radius = c(0.2, 0.2, 0.2, 0.2))

arrow_data <- data.frame(
  x_start = data$x[-length(data$x)],
  y_start = data$y[-length(data$y)],
  x_end = data$x[-1],
  y_end = data$y[-1],
  radius_start = data$radius[-length(data$radius)],
  radius_end = data$radius[-1])

arrow_data[3,c(1:2)] <- c(2,2)

# Adjust the start and end of the arrows to touch the circle's border
arrow_data$adjusted_x_start <- (arrow_data$x_start + arrow_data$radius_start *
  (arrow_data$x_end - arrow_data$x_start) / sqrt((arrow_data$x_end - arrow_data$x_start)^2 + (arrow_data$y_end - arrow_data$y_start)^2))

arrow_data$adjusted_y_start <- (arrow_data$y_start + arrow_data$radius_start *
  (arrow_data$y_end - arrow_data$y_start) / sqrt((arrow_data$x_end - arrow_data$x_start)^2 + (arrow_data$y_end - arrow_data$y_start)^2))

arrow_data$adjusted_x_end <- (arrow_data$x_end - arrow_data$radius_end *
  (arrow_data$x_end - arrow_data$x_start) / sqrt((arrow_data$x_end - arrow_data$x_start)^2 + (arrow_data$y_end - arrow_data$y_start)^2))

arrow_data$adjusted_y_end <- (arrow_data$y_end - arrow_data$radius_end *
  (arrow_data$y_end - arrow_data$y_start) / sqrt((arrow_data$x_end - arrow_data$x_start)^2 + (arrow_data$y_end - arrow_data$y_start)^2))

arrow_data$adjusted_y_start <-c(3.6, 1.6, 1.6)
arrow_data$adjusted_y_end   <-c(2.4, 0.62, 0.62)
arrow_data$adjusted_x_end   <-c(2, 0.72, 3.27)

edge_labels <- data.frame(
  x = (arrow_data$adjusted_x_start + arrow_data$adjusted_x_end)/2,
  y = (arrow_data$adjusted_y_start + arrow_data$adjusted_y_end)/2,
  label = c(
    "italic(f)[B~'|'~A](italic(x)[A]~','~italic(theta)[B~'|'~A])",
    "italic(f)[C~'|'~B](italic(x)[B]~','~italic(theta)[C~'|'~B])",
    "italic(f)[D~'|'~B](italic(x)[B]~','~italic(theta)[D~'|'~B])"
  )
)
edge_labels$x <-c(2, 1.1, 2.90)

fig_graph <- ggplot() +
  geom_point(data = data, aes(x = x, y = y), size = 14, shape = 21, fill = "white") +
  scale_size_continuous(range = c(1,15)) +
  geom_segment(data = arrow_data,
               aes(x = adjusted_x_start, y = adjusted_y_start, xend = adjusted_x_end, yend = adjusted_y_end),
               arrow = arrow(type = "closed", length = unit(.015, "npc")), color = "black", size = 0.5, lwd=0.2) +
  annotate("text",x = data$x,y = data$y,
  size = 4, parse = TRUE, label = paste0('italic(x)[', c("A", "B", "C", "D"), ']'))+
  geom_label(
    data = edge_labels,
    aes(x = x, y = y, label = label),
    parse = TRUE,
    fill = "white",
    size = 3,
    linewidth = 0.2
  ) +
  theme_void(base_family = "serif") +
  theme(legend.position = 'none') +
  xlim(0, 4) +
  ylim(0, 4.5) +
  theme(aspect.ratio = 1)
  
  zlist <- list()
set.seed(123)
for(z in 1:3){
x     <- data.frame(x=seq(-3,3,0.01), y=dnorm(seq(-3,3,0.01), 0, 1))
df    <- data.frame(y=rnorm(10), s=rgamma(10, 0.8^2/0.3^2, 0.8/0.3^2))
df$ll <- df$y-df$s
df$ul <- df$y+df$s

params <- abmeta(df$y, df$s, prior_mu = 0, prior_mu_se = 1)
y <- data.frame(x=seq(-3,3,0.01), y=dnorm(seq(-3,3,0.01), params[1], params[2]))
y <- y/max(y)

if(z!=1){xt_prior <- -1.5; yt_prior  <- 0.52
xt_like <- 0.4; yt_like <- 0.5
xt_post <- -0.32; yt_post <-  -0.25}else{
  xt_prior <- 1; yt_prior  <- 0.5
  xt_like <- 0.8; yt_like <- 0.49
  xt_post <- -0.5; yt_post <-  -0.2
}

zlist[[z]] <- cowplot::plot_grid(

ggplot(x, aes(x, y))+ylim(0,.6)+
  geom_line(lwd=.2)+theme_void(base_family = "serif")+
  annotate("text", x=xt_prior, y=yt_prior, size=2.5,
           label=expression("Prior:"~italic(P)(italic(theta)[B~ "|" ~A]~"|"~italic(I)))),

ggplot(df, aes(1:nrow(df), y=y))+
  geom_point(size=.6)+coord_flip()+theme_void(base_family = "serif")+
  geom_linerange(data=df, aes(ymin=ll, ymax=ul), lwd=.3),

ggplot(data.frame(x=c(0,1), y=c(0,1)),aes(x,y))+theme_void(base_family = "serif")+geom_point(col="white")+
  annotate("text", x=xt_like, y=yt_like, size=2.5, label=expression("Likelihood:"~italic(P)(italic(x)~"|"~italic(theta)[i], ..., italic(theta)[m]))),

ggplot(y, aes(x, y))+ylim(-1, max(y$y))+
        geom_line(lwd=.2)+theme_void(base_family = "serif")+annotate("text", x=xt_post, y=yt_post, size=2.5, label=expression("Posterior:"~italic(P)(italic(theta)[B~ "|" ~A]~"|"~italic(x), italic(I)))),

ncol=1, rel_heights = c(0.25, 0.3, 0.1, 0.35))}

ggplot(data.frame(a=1)) + xlim(0, 120) + ylim(0, 120) +
  annotation_custom(ggplotGrob(zlist[[2]]), xmin = 0, xmax = 40, ymin = 70, ymax = 120)+
  annotation_custom(ggplotGrob(fig_graph), xmin = 0, xmax = 160, ymin = -5, ymax = 110)+
  geom_path(data = as.data.frame(bezier(t = 0:100/100, p = list(x = c(23,23,68,68),
                                                                y = c(77, 77, 70, 70)))),
            aes(x = V1, y = V2), size = .8, col="black", lty=3, lwd=0.2, arrow = arrow(length = unit(.012, "npc"), type = "closed"))+
  theme_void(base_family = "serif")

```

Constructing a PPMN therefore requires two related specifications: the directed network structure and the parameterized edge-functions associated with each incoming edge. The following section shows how these components are defined in the package.

## Specifying network structure

We begin by defining each variable ($X = \{ x_1, ..., x_n \}$) that constitute the network and the directed dependencies among them. Each non-root vertex must be associated with a edge-function that maps its parent variable $x_p$ to a child vertex $x_c$. A edge-function is expressed as 

$$x_c=f_{c|p}(x_p, \theta_{c|p})$$ 
with it corresponding parameter $\theta_{c|p}$. An example could be the relations between Total Phosphorus (TP) to Chlorophyll-a (Chl). We fitted for example three models A-D and found one intercept (b0) and slope (b1) in two separated studies (this is given in `example4` below). The function is formally expressed as.
$$
x_{Chl}=f_{Chl|TP}(x_{TP}, \theta_{Chl|TP}) 
$$
However, the actual function is a linear equation where both the dependent and independent variable are log transformed. The expression of the function is to describe its relation and content. However, to the observant reader the fitted model is actually $log(y)=b0+b1*log(x)$ and describes the *deterministic* part of a GLM with log-link where the independent variable is log-transformed. Therefore, the expression above would in its predictive form be expressed as given below.
$$
x_{Chl}=f_{Chl|TP}(x_{TP}, \theta_{Chl|TP}) = exp\{\beta_0 + \beta_1 \cdot Log(x_{TP})\}
$$
We only needs to fit models to collect the `estimate` and `error` and indicate the functional relation between the dependent and independent variable(s) and how the independent variables have been transformed. As an example see the data frame (`example4`) with the following columns  below. In this data frame the columns `function`, `transformation`, `edge` `parameter`, `estimate` and `error` are required. The column `estimate` contains all numeric values of all intercept and slopes and the column `error` its corresponding standard error. We need to indicate to what `parameter` the `estimate` and `error` belong to. The parameter are b0 = $\beta_0$ and b1 = $\beta_1$, which in layman terms are called the intercept and slope, respectively. The column `edge` indicates the formula you would use in a standard `lm` or `glm` and is used to connect the dependent and independent variables. The column `transformation` indicates how all independent variables are transformed. It is applied to all independent variable and cannot be applied to one. The column `function` is important, it tells the code how we want to connect (what type of shape) the dependent and independent variables. The PPMN can apply different functions and used different types of parameters, that are not only intercepts or slopes and there can be functions added. Later we will come back to the functions currently available in the library. In the `example4` the columns `Source`, `Authors`, `Comment` could be omitted, but are displayed to show a table we would standard compile during a meta-analysis.

```{r, examp_1_ppmn_a, warning=F}

#load example
data("example4")

#display head
head(example4)

```

Each of the fitted models (`Source` = `A, B, C, D, E`) was a GLM, with log-link and log-transformed independent variable. It is also possible to only include an intercept or slope. To indicated the type of *deterministic* part of the GLM was used, the column `function` is indicated as *log*. Since the independent variable was log-transformed we denote *log* in the column `transformed`. If a log-transformation was not possible a square-root transformation *sqrt* can be used. If no transformation has been applied then the cell should remain empty. Other transformations can be added to the library on request. 

## Build PPMN function

We have a meta-analytic data frame (`example4`) with all columns and want to synthesize everything into a PPMN (which is at the moment only a single edge). We need to connect the data frame with information to the desired network structure and for this the `build_ppmn` function is used. This function builds the network structure and performs meta-analysis by connecting the columns `function`, `edge` `transformation` in the data frame (`example4`) to this structure. To specify each unique link we need to create a formula as given below. That formula should indicates the function `fun` *log* and the `edge` *Chl~TP* and type of transformation `trans` *log*. If no transformation is applied the argument `trans` can be omitted from the formula.

```{r, examp_1_ppmn_b, warning=F}

#We need to create a formula
formula=list(c(fun="log", edge="Chl~TP", trans="log"))

```

The PPMN however aims at maximizing information use. Therefore it uses a (empirical) Bayesian Meta-analysis in which the priors maximize the information to which a model cannot be fitted. Both the likelihood and prior in this model are assumed to approximate a Gaussian behavior. Therefore, the estimated parameters are specified via the mean `mu` ($\mu$) and standard error `se`. The prior for `mu` of each parameter b0, b1, b..., bn in an edge-function need tos be indicated with `prior_mu_b...` the prior for the `se` needs to be indicated with  `prior_se_b...`. If no prior is given the `mu` this defaults to 0 and if no prior is given for the `se` this defaults to 1000. Additionally, it is possible to truncate (cutoff) the distribution, restricting the outcome to only a part of the parameter space. For this we the lower truncation is indicated by `trunc_a_b...` and upper truncation > `trunc_b_b...`.

```{r, examp_1_ppmn_c, warning=F}
formula=list(c(fun="log", edge="Chl~TP", trans="log",
               prior_mu_b0=-0.5, prior_se_b0=0.5,
               prior_mu_b1=1, prior_se_b1=0.5, trunc_a_b1=0))

```

In our example a prior was already know the intercept falls around $-0.5$ and within the interval range of $~2.5\% = -0.5 - 2 \cdot 0.5=-1.5$ and $~97.5\% = -0.5 + 2 \cdot 0.5=0.5$. For the slope around $1$ and within the interval range of $~2.5\% = 1 - 2 \cdot 0.5=0$ and $~97.5\% = 1 + 2 \cdot 0.5=2$ we also assume the slope will definitely be >0 and therefore truncate the lower boundary (`trunc_a_b1`) at 0. This formula represents a single edge-function from TP to Chl. 

At this moment we have all ingredients (the formula and data) to build a foundational PPMN. To construct the foundational PPMN, both the formula and data need to be introduced into the `build_ppmn` function. This is shown below.

```{r, examp_1_ppmn_d, warning=F}

foundational_ppmn <- build_ppmn(formula = formula, data = example4)

```

After building the foundational PPMN, the single edge can be display in a DAG, 

```{r, examp_1_ppmn_e, warning=F}

foundational_ppmn$Plot_as_DAG

```

or as a Directed Bipartite Network, where the boxes represent the edge-functions and the variables the circles.

```{r, examp_1_ppmn_f, warning=F}

foundational_ppmn$Plot_as_BIP

```

A slightly more complex example would introduce use to a more complex edge-function and network. For example, we wish to regress *Chl~TN+TP* and use Chl to predict phytoplankton taxa richness of  *Taxa~Chl*. To bring a last formal expression of this network both edge-functions are described below.

$$
f_{Chl|TN, TP}(.) = f_{Chl|TN, TP}(x_{TN}, x_{TP}, \theta_{Chl|TN,TP,1},  \theta_{Chl|TN,TP,2}) = exp\{\beta_0 + \beta_1 \cdot Log(x_{TN}) + \beta_1 \cdot Log(x_{TP})\}
\\
f_{Taxa|Chl}(.)=f_{Taxa|Chl}(x_{Chl}, \theta_{Taxa|Chl})=exp(\beta_0 + \beta_1 \cdot Log(x_{Chl}))
$$
Both edge function are then combined to form a network expressed as below

$$
f_{Taxa|Chl}(f_{Chl|TN, TP}(x_{TN}, x_{TP}, \theta_{Chl|TN,TP,1},  \theta_{Chl|TN,TP,2}), \theta_{Taxa|Chl})
$$

The formula should then be specified as below. A minor addition is the inclusion of two priors for the edge *Taxa~Chl*. Assume there are two individuals authors/managers or methods to produce the priors then both sources of information are sources of uncertainty may be both be added. The meta-analysis incorporates this via Bayesian Model Averaging (BMA). Directly after building the foundational PPMN we choose to display the graphical representation as BIP.

```{r, examp_2_ppmn_a, warning=F}

data(example5)

head(example5)

#Create a formula
formula=list(c(fun="log", edge="Chl~TN+TP", trans="log",
               prior_mu_b0=-0.5, prior_se_b0=0.5,
               prior_mu_b1=0.5, prior_se_b1=0.5, trunc_a_b1=0,
               prior_mu_b2=1, prior_se_b2=0.5, trunc_a_b2=0),
             
             c(fun="log", edge="Taxa~Chl", trans="log",
               prior_mu_b0=0, prior_se_b0=5,
               prior_mu_b1=c(0.3, 0.5), prior_se_b1=c(0.5, 1), trunc_a_b1=0)
             
             )

#Build the foundational PPMN
foundational_ppmn <- build_ppmn(formula = formula, data = example5, 
                                node_width = 0.06,
                                fun_width = 0.1)

#Display the directed Bipartite graph
foundational_ppmn$Plot_as_BIP

```

## Marginal PPMN function

The network structure shows only one view of the information we incorporated. As with the results of Kaijser et al. (2025) we might be interested in viewing the association or the marginal edge-function itself an added variable plot. To do so an example data frame is created as given below. If we wish to omitted the points as we do not actually have real data then set the argument `alpha` to 0.

```{r, examp_2_ppmn_b,fig.width=8, fig.height=6, echo=F, warning=F}

library(cowplot)

df <- data.frame(TN=c(0.1, 1.25, 3.2, 2.23, 5), 
                 TP=c(5, 11, 27, 42, 83), 
                 Chl=c(0.2, 21, 30, 48, 92),
                 Taxa=c(4, 3, 9, 14, 25))

marginal_fig_Chl  <- marginal_ppmn(foundational_ppmn, response = "Chl", data =  df)
marginal_fig_Taxa <- marginal_ppmn(foundational_ppmn, response = "Taxa", data =  df)

cowplot::plot_grid(marginal_fig_Chl[[1]], #First object is TN
                   marginal_fig_Chl[[2]], #Second object is TP
                   marginal_fig_Taxa[[1]],#This is Chl 
                   ncol=1, labels = "AUTO")

```

## Predictive fit function

```{r, examp_2_ppmn_c, fig.width=8, fig.height=8, echo=F, warning=F}

pred_performance <- pred_fit(object = foundational_ppmn, new_data = df)

cowplot::plot_grid(pred_performance$plots$Stdz, #Predictive precision
                   pred_performance$plots$Balance, #Predictive balance
                   pred_performance$plots$Correlation, #Predictive correlation
                   ncol=1, labels = "AUTO")

```
